Benchmarking R packages for calculation of persistent homology

### 2020-12-28 tdabench-paper/tdabench.Rmd
added memory benchmarks

---
title: Benchmarking R packages for calculation of persistent homology
author:
  - name: Eashwar V. Somasundaram
    affiliation: School of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44106, United States
    email: eashwar.somasundaram@case.edu
  - name: Shael E. Brown
    affiliation: Department of Quantitative Life Sciences
    address:
    - McGill University
    - Montreal, QC H3A 0G4, Canada
    email: shael.brown@mail.mcgill.ca
  - name: Adam Litzler
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email:  adli4611@colorado.edu
  - name: Jacob G. Scott
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email: ScottJ10@ccf.org
  - name: Raoul R. Wadhwa
    affiliation: Cleveland Clinic Lerner College of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44195, United States
    email: raoul.wadhwa@case.edu
abstract: >
  Several persistent homology software libraries have been implemented in R. Specifically, the Dionysus, GUDHI, and Ripser libraries have been wrapped in R by the TDA and TDAstats packages on CRAN. They provide Vietoris-Rips and alpha complex functionality. These software represent powerful analysis tools that are currently computationally expensive and, to our knowledge, have not been formally benchmarked in R. Here, we compare the 2 R packages and the 3 underlying libraries for runtime and memory. We find that datasets with less than 3 dimensions can be evaluated with persistent homology fastest by the GUDHI library in the TDA package. For larger dimensional datasets, the Ripser library in the TDAstats package is best. Analysis of runtime and memory growth is also included in this report.
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
---

\section{Introduction}

@@

\section{Mathematics of Persistent Homology}
<!--change circles to more generalizable object, e.g. n-sphere or n-ball-->
An $n$-dimensional simplex is the convex hull of $n+1$ points in a Euclidean space.
More intuitively, an $n$-dimensional simplex is the simplest $n$-dimensional object (e.g. a 0-simplex is a point, a 1-simplex is a line, a 2-simplex is a triangle, 3-simplex is a tetrahedron).
These simplices can be glued together on common sub-simplices to form a simplicial complex (e.g. two triangles sharing a common side).
In a simplicial complex, topological features will arise that can be characterized by Betti numbers. 
Each Betti number, denoted by $B_k$, $k$ counts the number of features in dimension $k$. 
$B_0$ counts the number of connected components, $B_1$ counts refer to loops, $B_2$ counts the number of voids, and so on \citep{phom-survey}.

@@

\section{Methods}

@@

\section{Results}

@@

Vietoris-Rips and alpha complices differ not only in runtime, but also in memory usage.
GUDHI conveniently allows measurement of the simplicial complex boundary matrix, a reasonable proxy for total memory usage in calculation of persistent homology.
Since Ripser and Dionysus do not have similar capabilities, they are excluded from memory benchmarks.
Vietoris-Rips complices show rapid growth in object size as point cloud size increases; in contrast, alpha complices consistently maintain low, linearly-growing memory use (Figure \ref{fig:mem1}).
The growth pattern differences mirror findings for runtime growth (Figure \ref{fig:ann}).
Interestingly, different factors determine memory usage in the two types of simplicial complex.
Whereas memory use in calculation of a Vietoris-Rips complex is dependent on point cloud size and feature dimension, memory use for an analogous calculation with an alpha complex is also dependent on the structure of the point cloud itself (Figure \ref{fig:mem2}).
Within the limited set of tested point cloud structures, the sphere requires the least memory.
Table \ref{tbl:limit} lists the maximum point cloud size for which persistent homology calculations could be completed for each library.

\begin{figure}
  \centering
  \includegraphics[height=2.5in]{../Figures/Final_Figures/fig5a.png}
  \caption{
    \textbf{Comparing memory usage of persistent homology calculation using a Vietoris-Rips versus an alpha complex.}
    Mean boundary matrix size, a proxy for memory usage, plotted against point cloud size varying by point cloud structure (facet) and simplicial complex type (color, shape).
    Data dimension and feature dimension were held constant at 3 and 2, respectively.
    Point cloud structures for which memory usage was calculated were 3-dimensional annulus (top-left), 3-dimensional sphere (top-right), torus (bottom-left), and 3-dimensional cube (bottom-right).
    Similar to runtime results, alpha complex calculation showed linear growth and Vietoris-Rips complex calculation showed polynomial growth (see Supplement for regression details).}
\label{fig:mem1}
\end{figure}

\begin{widefigure}
  \centering
  \includegraphics[height=2.5in]{../Figures/Final_Figures/fig5t.png}
  \caption{
    \textbf{Factors affecting boundary matrix size in Vietoris-Rips and alpha complices.}
    Mean boundary matrix size plotted against point cloud size varying by point cloud structure (right, faceted), simplicial complex type (left, right), and feature dimension (color).
    Whereas the boundary matrix for a Vietoris-Rips complex (left) is independent of point cloud structure, the same is not true for an alpha complex (right, faceted).
    For the Vietoris-Rips complex, the decreasing number of data points plotted for successive feature dimensions reflects the computational requirements of increasing point cloud size.}
\label{fig:mem2}
\end{widefigure}

\begin{table}
\centering
\small
\begin{tabular}{|c|c|c|c|c|}
\hline
Feature & Ripser & Dionysus & GUDHI Rips & GUDHI Alpha \\
Dimension & Point Cloud Limit & Point Cloud Limit & Point Cloud Limit & Point Cloud Limit\\
\hline
1 & Not Reached & Not Reached & Not Reached & Not Reached \\ 
2 & Not Reached & 200 points & 280 points & Not Reached \\ 
3 & 300 points & 100 points & 125 points & Could Not Compute \\
4 & 100 points & 50 points & 75 points & Could Not Compute \\ 
\hline
\end{tabular}
\caption{During the benchmarking, certain libraries were not able to calculate persistent homology for sufficiently complex point clouds. 
Limiting factors were point cloud number and feature dimension calculation. 
For Vietoris-Rips complices, Ripser had the best performing capacity among the three libraries. 
Alpha complexes were not computable for any point cloud greater than 3 dimensions, so dimensional features 3 and higher could not be computed.}
\label{tbl:limit}
\end{table}

In real life scenarios, practitioners likely care most about comparative performance of the fastest Vietoris-Rips complex library and the fastest alpha complex library.
Figure \ref{fig:fin} illustrates this comparison, showing that for 3-dimensional point clouds, GUDHI's alpha complex library outpaces Ripser's Vietoris-Rips functionality.

\section{Discussion}
As persistent homology calculations continue to become a more popular tool to analyze complex multidimensional data, it will be important to understand from a computational perspective which method to use. 
In this paper, we examined two forms of persistent homology complexes: Vietoris-Rips and alpha complexes. 
Both algorithms describe topological features through the generation of simplicial complexes. 
The advantage in saving computational time by choosing a particular algorithm depends on point cloud characteristics.

>>>>

---
title: Benchmarking R packages for calculation of persistent homology
author:
  - name: Eashwar V. Somasundaram
    affiliation: School of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44106, United States
    email: eashwar.somasundaram@case.edu
  - name: Shael E. Brown
    affiliation: Department of Quantitative Life Sciences
    address:
    - McGill University
    - Montreal, QC H3A 0G4, Canada
    email: shael.brown@mail.mcgill.ca
  - name: Adam Litzler
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email:  adli4611@colorado.edu
  - name: Jacob G. Scott
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email: ScottJ10@ccf.org
  - name: Raoul R. Wadhwa
    affiliation: Cleveland Clinic Lerner College of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44195, United States
    email: raoul.wadhwa@case.edu
abstract: >
  Several persistent homology software libraries have been implemented in R. Specifically, the Dionysus, GUDHI, and Ripser libraries have been wrapped in R by the TDA and TDAstats packages on CRAN. They provide Vietoris-Rips and alpha complex functionality. These software represent powerful analysis tools that are currently computationally expensive and, to our knowledge, have not been formally benchmarked in R. Here, we provide an analysis of runtime and memory growth for the 2 R packages and the 3 underlying libraries. We find that datasets with less than 3 dimensions can be evaluated with persistent homology fastest by the GUDHI library in the TDA package. For higher dimensional datasets, the Ripser library in the TDAstats package is best. 
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
---

```{r setup, echo = FALSE}
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(broom))
```

# Introduction

@@

# Mathematics of Persistent Homology
<!--change circles to more generalizable object, e.g. n-sphere or n-ball-->
An $n$-dimensional simplex is the convex hull of $n+1$ points in a Euclidean space.
More intuitively, an $n$-dimensional simplex is the simplest $n$-dimensional object (e.g. a 0-simplex is a point, a 1-simplex is a line, a 2-simplex is a triangle, 3-simplex is a tetrahedron).
These simplices can be glued together on common sub-simplices to form a simplicial complex (e.g. two triangles sharing a common side).
In a simplicial complex, topological features will arise that can be characterized by Betti numbers. 
Each Betti number, denoted by $B_k$, $k$ counts the number of features in dimension $k$. 
$B_0$ counts the number of connected components, $B_1$ counts refer to loops, $B_2$ counts the number of voids, and so on \citep{phom-survey}.

@@

# Methods

@@

# Results

@@

```{r powerlaw, echo=FALSE}
# input data
df_mem <- read.csv("../memory-bench/memory-use.csv") %>%
  mutate(Memory = Memory / 1024)

# setup df
pl_exponents <- expand.grid(c("TDAstats", "GUDHI", "Dionysus"),
                            2:4,
                            c("circle", "annulus", "torus", "uniform"))

# return power law exponent from data
calc_pl_exp <- function(all_data, engine, dim, shape) {
  # only >= 100 points
  all_data <- all_data %>%
    filter(NumPoints >= 100 &
           Engine == engine &
           Dimension == dim &
           DataType == shape) %>%
    mutate(LogPts = log10(NumPoints),
           LogMem = log10(Memory))

  lm_mod <- lm(LogMem ~ LogPts, data = all_data)
  tidy_mod <- tidy(lm_mod)

  return(tidy_mod$estimate[2])
}
```

In addition to runtime differences, the three Vietoris-Rips homology engines differ in memory use.
After crossing a threshold between 50 and 100 points, all three engines appeared to follow power law growth, with a linear trend on log-log plots (Figure \ref{fig:mem}).
However, for all combinations of point cloud dimension and shape, TDAstats used the least memory and Dionysus used the most, with TDAstats also growing with the smallest power law exponent as the number of points increased.
As an example, for 3-dimensional annulus point clouds, GUDHI memory use grew with a power law exponent of
`r calc_pl_exp(df_mem, "GUDHI", 3, "annulus") %>% signif(digits = 2)`,
Dionysus use grew with an exponent of
`r calc_pl_exp(df_mem, "Dionysus", 3, "annulus") %>% signif(digits = 2)`,
and TDAstats use grew with an exponent of
`r calc_pl_exp(df_mem, "TDAstats", 3, "annulus") %>% signif(digits = 2)`.
The entire table of exponents can be found in the Supplement (GitHub), but Figure \ref{fig:mem} clearly illustrates that TDAstats was, for all tested cases, the optimal Vietoris-Rips engine in the context of memory use.

```{r memory, echo=FALSE, fig.cap="\\label{fig:mem}\\textbf{Comparing memory use of Vietoris-Rips persistent homology engines.} Each column title corresponds to point cloud dimension; each row title lists point cloud shape; each persistent homology engine is represented by points of a distinct shape and color. For point clouds containing more than 50 points, there appears to be a linear trend on the log-log axes. Data for 2- and 4-dimensional tori were not collected because a torus does not trivially generalize to dimensions other than 3. Missing points for GUDHI and Dionysus in the 4-dim plots indicate that persistent homology calculation was terminated since memory requirement exceeded available RAM (32 GB)."}
df_mem <- df_mem %>%
  mutate(Dimension = factor(Dimension, levels = 2:4, labels = c("2-dim", "3-dim", "4-dim")))

ggplot(df_mem, aes(x = NumPoints, y = Memory, color = Engine, shape = Engine)) +
  geom_point() +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  facet_grid(DataType ~ Dimension) +
  xlab("Number of points") +
  ylab("Memory (GB)") +
  theme_bw()
```

# Discussion

As persistent homology calculations continue to become a more popular tool to analyze complex multidimensional data, it will be important to understand from a computational perspective which method to use. 
In this paper, we examined two forms of persistent homology complexes: Vietoris-Rips and alpha complexes. 
Both algorithms describe topological features through the generation of simplicial complexes. 
The advantage in saving computational time by choosing a particular algorithm depends on point cloud characteristics.
